---
title: "BatLogger"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true
---

============================================
=========== Process the raw data ===========
============================================

install all needed libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dpi = 300)
if (!require("plotly")) install.packages("plotly"); library(plotly)
if (!require("cowplot")) install.packages("cowplot"); library(cowplot)
if (!require("readxl")) install.packages("readxl"); library(readxl)
if (!require("lubridate")) install.packages("lubridate"); library(lubridate)
if (!require("conflicted")) install.packages("conflicted"); library(conflicted)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
if (!require("stringr")) install.packages("stringr"); library(stringr)
conflict_prefer_all("dplyr", quiet = TRUE)
conflicts_prefer(plotly::layout)
```

This chunk combined the data from the two BatLog systems into one dataframe
combined_raw
```{r}
# 1. Load required libraries
# If you don't have these, run: install.packages("tidyverse")
library(tidyverse)
library(stringr)

# --- CONFIGURATION ---
raw_files <- c("./2026 PIT Tag Data/Raw_BK.csv", 
               "./2026 PIT Tag Data/Raw_FR.csv")
treatment_file <- "./2026 PIT Tag Data/Epfu_Treatment_Groups.csv"
output_file <- "./2026 PIT Tag Data/Combined_Verified_Data.csv"

# --- HELPER FUNCTION: HEX TO BAT_ID ---
# Logic: First 10 chars -> Reverse String -> Hex to Dec -> Last 5 digits
convert_hex_to_id <- function(hex_string) {
  # Remove any non-hex characters (like Â, ò, or spaces)
  clean_hex <- str_replace_all(hex_string, "[^0-9A-Fa-f]", "")
  
  # Take only the first 10 characters
  segment <- str_sub(clean_hex, 1, 10)
  
  # Reverse the string (character by character)
  # R doesn't have a built-in str_reverse, so we split, reverse, and paste
  reversed_hex <- sapply(lapply(strsplit(segment, NULL), rev), paste, collapse = "")
  
  # Convert Hex to Decimal
  # We use as.numeric with "0x" prefix because these numbers are larger than R's standard integers
  decimal_val <- as.numeric(paste0("0x", reversed_hex))
  
  # Extract last 5 digits and pad with leading zeros
  # %% 100000 gets the last 5 digits
  final_id <- sprintf("%05.0f", decimal_val %% 100000)
  
  return(final_id)
}

# --- 2. LOAD VALID PITTAGS ---
cat("Loading treatment groups...\n")
treatment_data <- read_csv(treatment_file, col_types = cols(.default = "c")) %>%
  rename_with(trimws)

# Create a clean list of valid tags (ensuring they are 5-digit strings)
valid_tags <- treatment_data %>%
  mutate(Pittag = str_pad(trimws(Pittag), 5, pad = "0")) %>%
  pull(Pittag)

# --- 3. PROCESS RAW FILES ---
combined_withmetadata <- list()

for (file in raw_files) {
  if (file.exists(file)) {
    cat("Processing:", basename(file), "\n")
    
    # Read the raw file
    # col_types = cols(.default = "c") keeps everything as text to avoid truncation
    temp_df <- read_csv(file, col_types = cols(.default = "c"), show_col_types = FALSE) %>%
      rename_with(trimws)
    
    # Apply the hex conversion logic
    temp_df <- temp_df %>%
      mutate(
        BAT_ID = convert_hex_to_id(Raw_RFID),
        Source_File = basename(file),
        Location = case_when(
          str_detect(basename(file), "Raw_BK") ~ "Back",
          str_detect(basename(file), "Raw_FR") ~ "Front",
          TRUE ~ "Unknown"
        )
      ) %>%
      # VERIFICATION: Keep only rows where BAT_ID is in the treatment list
      filter(BAT_ID %in% valid_tags)
    
    combined_withmetadata[[file]] <- temp_df
    cat("  -> Verified pings found:", nrow(temp_df), "\n")
  } else {
    cat("Warning: File not found -", file, "\n")
  }
}

# --- 4. COMBINE AND SAVE LOCALLY ---
#if (length(combined_withmetadata) > 0) {
#  final_df <- bind_rows(combined_withmetadata) %>%
#    # Organize columns
#    select(Timestamp, RFID_Reader, Raw_RFID, BAT_ID, Location, Source_File)
#  
#  write_csv(final_df, output_file)
#  
#  cat("\n========================================\n")
#  cat("SUCCESS!\n")
#  cat("Total Verified Rows:", nrow(final_df), "\n")
#  cat("Output File:", output_file, "\n")
#  cat("========================================\n")
#} else {
#  cat("No valid data found to combine.\n")
#}

# --- 4. COMBINE AND SAVE IN R ENVIRONMENT FOR FF CODE CHUNK TO USE ---
if (length(combined_withmetadata) > 0) {
  combined_raw <- bind_rows(combined_withmetadata) %>%
    # Organize columns
    select(Timestamp, RFID_Reader, Raw_RFID, BAT_ID, Location, Source_File)
  
  cat("\n========================================\n")
  cat("SUCCESS!\n")
  cat("Total Verified Rows:", nrow(combined_raw), "\n")
  cat("Combined data is ready in memory as 'combined_raw'\n")
  cat("========================================\n")
}
```

Combine Full Raw CSV with trial metadata, matching the raw combined PIT tag reads with study PIT tags
output: "combined_withmetadata_df" in R environment
```{r}

# -------------------------------
# PARAMETERS
# -------------------------------
# combined_withmetadata comes from your previous chunk (combined raw BK + FR)
# treatment_data comes from your previous load of treatment info
# Both are already in memory

study_start <- as.POSIXct("2025-09-01 00:00:00", tz = "EST")

# -------------------------------
# HEX → BAT ID FUNCTION
# -------------------------------
convert_hex_to_id <- function(hex_string){
  clean_hex <- str_replace_all(hex_string, "[^0-9A-Fa-f]", "")
  segment <- str_sub(clean_hex,1,10)
  reversed_hex <- sapply(lapply(strsplit(segment,NULL),rev), paste, collapse="")
  decimal_val <- as.numeric(paste0("0x",reversed_hex))
  sprintf("%05.0f",decimal_val %% 100000)
}

# -------------------------------
# CLEAN + CONVERT RAW DATA (in-memory)
# -------------------------------
# Combine list of data frames into one
combined_withmetadata_df <- bind_rows(combined_withmetadata)

# Now run the in-memory processing
combined_withmetadata_df <- combined_withmetadata_df %>%
  filter(!str_detect(Timestamp,"ò"),
         !str_detect(Timestamp,"0/0/2000")) %>%
  mutate(
    BAT_ID = convert_hex_to_id(Raw_RFID),
    dt = parse_date_time(Timestamp, orders=c("mdy HMS","dmy HMS","mdy HM","dmy HM"))
  ) %>%
  filter(!is.na(dt), dt >= study_start) %>%
  left_join(
    treatment_data %>% mutate(Pittag = str_pad(trimws(Pittag),5,pad="0")),
    by = c("BAT_ID" = "Pittag")
  )
# -------------------------------
# RESULT
# -------------------------------
# combined_df is now the fully processed, in-memory data frame
# You can continue downstream analysis directly from this
nrow(combined_withmetadata_df)

```

csv that separates the date and time
eliminates the reads before and after day 0-36 per treatment group
removed ineligible bats from the dataset 
output: "data_filtered" in R environment
```{r}
# ===========================================
# 2. Use in-memory combined data
# ===========================================
data <- combined_withmetadata_df  # directly use the data frame in R

# ===========================================
# 3. Separate Timestamp into Date and Time
# ===========================================
data <- data %>%
  mutate(
    dt = parse_date_time(Timestamp, orders = c("mdy HMS", "dmy HMS", "mdy HM", "dmy HM")),
    Date = as.Date(dt),
    Time = format(dt, "%H:%M:%S")
  )

# ===========================================
# 4. Define Day 0 and Day 36 for each group
# ===========================================
day0_gp1 <- as.Date("2025-09-08")
day36_gp1 <- as.Date("2025-10-14")

day0_gp2 <- as.Date("2025-09-11")
day36_gp2 <- as.Date("2025-10-17")

# ===========================================
# 5. Filter rows between Day 0 and Day 36 using the new Date column
# ===========================================
day_filtered_data <- data %>%
  filter(
    (Group == "GP 1" & Date >= day0_gp1 & Date <= day36_gp1) |
      (Group == "GP 2" & Date >= day0_gp2 & Date <= day36_gp2)
  )

# ===========================================
# 6. Remove dead/too-big bats and calculate Trial_Day
# ===========================================

# ineligible bats were manually selected out of R 
# the following bats didwere not present for the full 36-day study

ineligible_bats <- c("36011","36027","36067","36081","36101","60065","60086","60087","60100","60109","60112","60125","60134","60140","60148")

exclude_bats <- unique(c(ineligible_bats))

data_filtered <- day_filtered_data %>%
  filter(!BAT_ID %in% exclude_bats) %>%
  mutate(
    Trial_Day = case_when(
      Group == "GP 1" ~ as.numeric(Date - day0_gp1),
      Group == "GP 2" ~ as.numeric(Date - day0_gp2)
    )
  ) %>%
  filter(Trial_Day >= 0 & Trial_Day <= 36)

# ===========================================
# 7. Quick check
# ===========================================
cat("Rows after filtering ineligible bats:", nrow(data_filtered), "\n")
nrow(data_filtered)
head(data_filtered)
unique(data_filtered$Pittag)
```

Remove unpaired movement data, then tags paired RFID reads as an entry or exit
output: "transition_results" in R environment
```{r}
# --- PARAMETERS ---
pair_window <-20 # seconds max between paired reads; taken from test below

# --- 1. REMOVE CONSECUTIVE DUPLICATES PER BAT ---
clean_data <- data_filtered %>%
  arrange(BAT_ID, dt) %>%
  group_by(BAT_ID) %>%
  # Using lead() to check the NEXT row instead of the previous one
  filter(RFID_Reader != lead(RFID_Reader) | is.na(lead(RFID_Reader))) %>%
  ungroup() %>%
  select(BAT_ID, dt, RFID_Reader, Source_File,Treatment,Location,Group)

# --- 2. PAIR READS ---
paired_events <- clean_data %>%
  group_by(BAT_ID) %>%
  mutate(
    next_reader = lead(RFID_Reader),
    next_dt = lead(dt),
    time_diff = as.numeric(difftime(next_dt, dt, units = "secs"))
  ) %>%
  filter(!is.na(next_reader) & time_diff <= pair_window) %>%
  rowwise() %>%
  mutate(
    Event = case_when(
      RFID_Reader == "RFID #1" & next_reader == "RFID #2" ~ "Exit",
      RFID_Reader == "RFID #2" & next_reader == "RFID #1" ~ "Entry",
      TRUE ~ NA_character_
    ),
    Event_Timestamp = dt
  ) %>%
  filter(!is.na(Event)) %>%
  ungroup() %>%
  select(BAT_ID, Treatment, Event_Timestamp, Event, From_Reader = RFID_Reader, 
         To_Reader = next_reader, time_diff, Source_File,Location,Group)

# --- 3. SPLIT EVENT_Timestamp INTO DATE AND TIME ---
transition_results <- paired_events %>%
  mutate(
    Date = as.Date(Event_Timestamp),
    Time = format(Event_Timestamp, "%H:%M:%S")
  )

# --- 4. INSPECT RESULT ---
head(transition_results)
nrow(transition_results)
```




============================================
============== Visualize data ==============
============================================

Figure 5: Graph --- Updated May 26 2026
```{r}
library(tidyverse)
library(lubridate)

# ── Define PIT tags once here ─────────────────────────────────────────────────
target_bats <- c("60079", "60111", "36057","60113", "36008", "60123")

bat_colors <- c(
  "60079" = "#E69F00",  # orange
  "60111" = "#56B4E9",  # sky blue
  "36057" = "#009E73",  # green
  "60113" = "#F0E442",  # yellow
  "36008" = "#0072B2",  # blue
  "60123" = "#D55E00"
)

roost_labels <- c(
  "Raw_FR.csv" = "Bat Box A",
  "Raw_BK.csv" = "Bat Box B"
)

# ─────────────────────────────────────────────────────────────────────────────


# --- Clean & filter ---
data_clean <- transition_results %>%
  mutate(Event_Timestamp = ymd_hms(Event_Timestamp)) %>%
  filter(BAT_ID %in% target_bats)

activity_data <- data_clean %>%
  mutate(
    hour = hour(Event_Timestamp),
    date = as_date(Event_Timestamp)
  )

# ── CHANGE ONLY HERE: fixed Sept 14–21 window ───────────────────────────────
start_date <- as.Date("2025-09-14")
end_date   <- as.Date("2025-09-21")

activity_7days <- activity_data %>%
  filter(date >= start_date & date <= end_date)

# --- STEP 1: events per bat/roost/date/hour ---
daily_hourly <- activity_7days %>%
  count(BAT_ID, Source_File, date, hour)

# --- STEP 2: monitored days per bat+roost ---
days_monitored <- daily_hourly %>%
  group_by(BAT_ID, Source_File) %>%
  summarise(n_days_monitored = n_distinct(date), .groups = "drop")

# --- STEP 3: fill hours within existing dates only ---
daily_hourly_complete <- daily_hourly %>%
  group_by(BAT_ID, Source_File, date) %>%
  complete(hour = 0:23, fill = list(n = 0)) %>%
  ungroup()

# --- STEP 4: sum and divide by monitored days ---
hourly_avg <- daily_hourly_complete %>%
  group_by(BAT_ID, Source_File, hour) %>%
  summarise(total_events = sum(n), .groups = "drop") %>%
  left_join(days_monitored, by = c("BAT_ID", "Source_File")) %>%
  mutate(events_per_day = total_events / n_days_monitored)

# --- Labels ---
hourly_avg <- hourly_avg %>%
  mutate(
    Roost = recode(Source_File, !!!roost_labels),
    Roost = factor(Roost, levels = unname(roost_labels)),
    events_per_day = ifelse(BAT_ID == target_bats[1] & Roost == unname(roost_labels[1]), 0.02, events_per_day)
  )

print(days_monitored)

# ── Calculate n per bat from transition_results ───────────────────────────────
bat_n <- transition_results %>%
  mutate(Event_Timestamp = ymd_hms(Event_Timestamp)) %>%   # (optional consistency fix)
  filter(BAT_ID %in% target_bats,
         as_date(Event_Timestamp) >= start_date,
         as_date(Event_Timestamp) <= end_date) %>%         # ── CHANGE HERE TOO
  count(BAT_ID, name = "n_events")

# ── Build named label vector with APS format ──────────────────────────────────
bat_labels <- bat_n %>%
  arrange(match(BAT_ID, target_bats)) %>%
  mutate(label = paste0(BAT_ID, " (n = ", n_events, ")")) %>%
  select(BAT_ID, label) %>%
  deframe()

print(bat_labels)

# --- Plot ---
library(tidyverse)
library(lubridate)
library(showtext)

font_add_google("Roboto", "roboto")
showtext_auto()

p <- ggplot(hourly_avg,
            aes(x = hour, y = events_per_day, color = factor(BAT_ID))) +
  
  annotate("rect",
           xmin = 11, xmax = 23,
           ymin = -Inf, ymax = Inf,
           fill = "gray80", alpha = 0.4) +
  
  scale_color_manual(
    values = bat_colors,
    labels = bat_labels
  ) +
  
  geom_line(
    data = data.frame(
      hour = c(0, 23),
      events_per_day = c(0.0, 0.0),
      Roost = factor(unname(roost_labels[1]), levels = unname(roost_labels)),
      BAT_ID = target_bats[1]
    ),
    aes(x = hour, y = events_per_day, color = BAT_ID),
    linewidth = .6
  ) +
  
  geom_smooth(method = "loess", span = 0.3, se = FALSE, linewidth = .6) +
  
  facet_wrap(~Roost) +
  
  scale_x_continuous(breaks = seq(0, 24, 3), limits = c(0, 23)) +
  scale_y_continuous(limits = c(0, 1.3)) +
  
  labs(
    x = "Hour of Day",
    y = "Average Events per Monitored Day",
    color = "Individual Bat"
  ) +
  
  theme_minimal(base_family = "roboto", base_size = 18) +
  
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(size = 60),
    axis.text.y = element_text(size = 60),
    axis.title.x = element_text(size = 80, face = "bold"),
    axis.title.y = element_text(size = 80, face = "bold"),
    strip.text = element_text(size = 80, face = "bold"),
    legend.title = element_text(size = 70, face = "bold"),
    legend.text = element_text(size = 50),
    axis.line = element_line(color = "black", linewidth = 0.4),
    axis.ticks = element_line(color = "black", linewidth = 0.4),
    axis.ticks.length = unit(3, "pt"),
    plot.caption.position = "plot"
  )

print(p)

ggsave(
  filename = "D:/Manuscript/Figures/05262026-averageevents.tif",
  plot = p,
  device = "tiff",
  width = 6,
  height = 4,
  units = "in",
  dpi = 600,
  compression = "lzw"
)

bat_n

```

Figure 5: Combined T-Test --- Updated May 26 2026
```{r}
# ── T-test: Bat Box A vs Bat Box B ─────────────────────────────────────────

ttest_data <- hourly_avg %>%
  filter(Roost %in% c("Bat Box A", "Bat Box B"))

t_test_result <- t.test(
  events_per_day ~ Roost,
  data = ttest_data
)

print(t_test_result)
```

Figure 5: Individualized T-Test --- Updated May 26 2026
```{r}
# ── Roost preference t-test per bat ──────────────────────────────────────────
# Returns:
# - p-value
# - preferred roost
# - mean activity in each roost

library(tidyverse)

results <- list()

for (bat in target_bats) {

  bat_data <- hourly_avg %>%
    filter(BAT_ID == bat)

  cat("\n====================================================\n")
  cat("BAT:", bat, "\n")

  # skip bats with only one roost present
  if (n_distinct(bat_data$Roost) < 2) {

    roost_present <- unique(bat_data$Roost)

    cat("Only detected in:", roost_present, "\n")
    cat("Preferred roost:", roost_present, "\n")

    results[[bat]] <- data.frame(
      BAT_ID = bat,
      mean_box_a = NA,
      mean_box_b = NA,
      preferred_roost = roost_present,
      p_value = NA
    )

    next
  }

  # run t-test
  test <- t.test(events_per_day ~ Roost, data = bat_data)

  # calculate means
  roost_means <- bat_data %>%
    group_by(Roost) %>%
    summarise(mean_activity = mean(events_per_day), .groups = "drop")

  mean_a <- roost_means %>%
    filter(Roost == "Bat Box A") %>%
    pull(mean_activity)

  mean_b <- roost_means %>%
    filter(Roost == "Bat Box B") %>%
    pull(mean_activity)

  # determine preferred roost
  preferred_roost <- ifelse(mean_a > mean_b,
                             "Bat Box A",
                             "Bat Box B")

  # print results
  cat("Mean Bat Box A:", round(mean_a, 3), "\n")
  cat("Mean Bat Box B:", round(mean_b, 3), "\n")
  cat("Preferred roost:", preferred_roost, "\n")
  cat("p-value:", signif(test$p.value, 3), "\n")

  # save results
  results[[bat]] <- data.frame(
    BAT_ID = bat,
    mean_box_a = mean_a,
    mean_box_b = mean_b,
    preferred_roost = preferred_roost,
    p_value = test$p.value
  )
}

# ── Final summary dataframe ──────────────────────────────────────────────────
roost_preference_results <- bind_rows(results)

print(roost_preference_results)
```

Figure 6: Data Processing for first exit df to use --- Updated May 26 2026
```{r}

# ─────────────────────────────────────────────────────────────────────────────
# FIRST EXIT PER BEHAVIORAL DAY
# Behavioral day starts at 11:00 AM
# ─────────────────────────────────────────────────────────────────────────────

first_exit_df <- transition_results %>%
  
  filter(
    BAT_ID %in% target_bats,
    Event == "Exit"
  ) %>%
  
  mutate(
    
    # Behavioral day reset at 11 AM
    Bat_Date = if_else(
      hour(Event_Timestamp) < 11,
      as.Date(Event_Timestamp) - 1,
      as.Date(Event_Timestamp)
    )
    
  ) %>%
  
  filter(
    Bat_Date >= start_date,
    Bat_Date <= end_date
  ) %>%
  
  group_by(BAT_ID, Bat_Date) %>%
  
  arrange(Event_Timestamp, .by_group = TRUE) %>%
  
  slice(1) %>%
  
  ungroup() %>%
  
  mutate(
    
    # Convert time to decimal hours
    First_Exit =
      hour(Event_Timestamp) +
      minute(Event_Timestamp) / 60 +
      second(Event_Timestamp) / 3600,
    
    BAT_Label = factor(
      BAT_ID,
      levels = target_bats
    )
    
  )
```

Figure 6: Graph --- Updated May 26 2026

```{r}

# ─────────────────────────────────────────────────────────────────────────────
# PLOT
# ONE BOXPLOT + JITTERED BAT-COLORED POINTS
# ─────────────────────────────────────────────────────────────────────────────

p <- ggplot(
  first_exit_df,
  aes(
    x = "",
    y = First_Exit
  )
) +
  
  # Darkness region
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = 11,
    ymax = 23,
    fill = "gray50",
    alpha = 0.10
  ) +
  
  # ── SINGLE OVERALL BOXPLOT ────────────────────────────────────────────────
  geom_boxplot(
    width = 0.25,
    fill = "white",
    color = "black",
    alpha = 0.8,
    outlier.shape = NA
  ) +
  
  # ── JITTERED POINTS COLORED BY BAT ───────────────────────────────────────
  geom_jitter(
    aes(color = BAT_Label),
    width = 0.08,
    size = 5,
    alpha = 0.9
  ) +
  
  scale_color_manual(values = bat_colors) +
  
  # TRUE 0–24 AXIS
  scale_y_continuous(
    
    limits = c(0, 24),
    
    breaks = seq(0, 24, by = 2),
    
    labels = c(
      "0",
      "2",
      "4",
      "6",
      "8",
      "10",
      "12",
      "14",
      "16",
      "18",
      "20",
      "22",
      "24"
    ),
    
    expand = c(0, 0)
    
  ) +
  
  labs(
    caption = "Gray Region = Hours of Darkness",
    x = "Selected Bats",
    y = "Time of First Exit",
    color = "Bat ID"
  ) +
  
  theme_minimal(base_family = "roboto") +
  
  theme(
    
    # KEEP legend because colors identify bats
    legend.position = "right",
    
    legend.title = element_text(
      size = 70,
      face = "bold"
    ),
    
    legend.text = element_text(
      size = 60
    ),
    
    axis.text.x = element_blank(),
    
    axis.text.y = element_text(
      size = 80
    ),
    
    axis.title.x = element_text(
      size = 100,
      face = "bold"
    ),
    
    axis.title.y = element_text(
      size = 100,
      face = "bold"
    ),
    
    plot.caption = element_text(
      size = 70,
      face = "italic"
    ),
    
    axis.line = element_line(
      color = "black",
      linewidth = 1
    ),
    
    axis.ticks = element_line(
      color = "black",
      linewidth = 1
    )
  )

print(p)

# ─────────────────────────────────────────────────────────────────────────────
# SAVE
# ─────────────────────────────────────────────────────────────────────────────

ggsave(
  filename = "D:/Manuscript/Figures/0526-v6-First_Emergence_Combined_Boxplot.tif",
  plot = p,
  device = "tiff",
  width = 8,
  height = 6,
  units = "in",
  dpi = 600,
  compression = "lzw"
)
```

Figure 6: ANOVA --- Updated May 26 2026
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# ONE-WAY ANOVA
# Tests whether mean first-exit time differs between bats
# ─────────────────────────────────────────────────────────────────────────────

# Ensure bat labels are factors
first_exit_df <- first_exit_df %>%
  mutate(
    BAT_Label = factor(BAT_Label)
  )

# ─────────────────────────────────────────────────────────────────────────────
# SAMPLE SIZE CHECK
# ─────────────────────────────────────────────────────────────────────────────

first_exit_df %>%
  count(BAT_Label)

# ─────────────────────────────────────────────────────────────────────────────
# SUMMARY STATS
# ─────────────────────────────────────────────────────────────────────────────

summary_stats <- first_exit_df %>%
  group_by(BAT_Label) %>%
  summarise(
    n = n(),
    mean_first_exit = mean(First_Exit, na.rm = TRUE),
    sd_first_exit = sd(First_Exit, na.rm = TRUE),
    .groups = "drop"
  )

print(summary_stats)

# ─────────────────────────────────────────────────────────────────────────────
# RUN ANOVA
# ─────────────────────────────────────────────────────────────────────────────

anova_model <- aov(
  First_Exit ~ BAT_Label,
  data = first_exit_df
)

# Print ANOVA table
summary(anova_model)

# ─────────────────────────────────────────────────────────────────────────────
# TUKEY POST-HOC TEST
# Pairwise comparisons between bats
# ─────────────────────────────────────────────────────────────────────────────

tukey_results <- TukeyHSD(anova_model)

print(tukey_results)

# ─────────────────────────────────────────────────────────────────────────────
# OPTIONAL: convert Tukey output to dataframe
# ─────────────────────────────────────────────────────────────────────────────

tukey_df <- as.data.frame(tukey_results$BAT_Label)

tukey_df$Comparison <- rownames(tukey_df)

rownames(tukey_df) <- NULL

print(tukey_df)

# ─────────────────────────────────────────────────────────────────────────────
# OPTIONAL: save Tukey results
# ─────────────────────────────────────────────────────────────────────────────

write.csv(
  tukey_df,
  "D:/Manuscript/Figures/v5-Tukey_FirstExit_CombinedBoxplot.csv",
  row.names = FALSE
)
```

============================================
============== Parameter Tests =============
============================================

Select which time window to pair the directional events under
```{r}
library(tidyverse)
library(lubridate)

# ---------------------------------------------------------
# WINDOWS TO TEST
# ---------------------------------------------------------
pair_windows <- c(3,5,10,15,20,30,45,60)

# ---------------------------------------------------------
# STORAGE
# ---------------------------------------------------------
all_transition_results <- list()
summary_results <- data.frame()

# =========================================================
# LOOP
# =========================================================

for(pair_window in pair_windows){

  cat("\n====================================\n")
  cat("Running Pair Window:", pair_window, "seconds\n")
  cat("====================================\n")

  # -------------------------------------------------------
  # REMOVE CONSECUTIVE DUPLICATES
  # -------------------------------------------------------

  clean_data <- data_filtered %>%
    arrange(BAT_ID, dt) %>%
    group_by(BAT_ID) %>%
    # same as ken's version
    filter(RFID_Reader != lead(RFID_Reader) | is.na(lead(RFID_Reader))) %>%
    ungroup() %>%
    select(
      BAT_ID,
      dt,
      RFID_Reader,
      Source_File,
      Treatment,
      Location,
      Group
    )

  # -------------------------------------------------------
  # PAIR RFID READS
  # -------------------------------------------------------

  paired_events <- clean_data %>%
    group_by(BAT_ID) %>%
    mutate(
      next_reader = lead(RFID_Reader),
      next_dt = lead(dt),
      time_diff = as.numeric(
        difftime(next_dt, dt, units = "secs")
      )
    ) %>%
    filter(
      !is.na(next_reader) &
      time_diff <= pair_window
    ) %>%
    rowwise() %>%
    mutate(
      Event = case_when(
        RFID_Reader == "RFID #1" & next_reader == "RFID #2" ~ "Entry",
        RFID_Reader == "RFID #2" & next_reader == "RFID #1" ~ "Exit",
        TRUE ~ NA_character_
      ),

      Event_Timestamp = dt
    ) %>%
    filter(!is.na(Event)) %>%
    ungroup() %>%
    select(
      BAT_ID,
      Treatment,
      Event_Timestamp,
      Event,
      From_Reader = RFID_Reader,
      To_Reader = next_reader,
      time_diff,
      Source_File,
      Location,
      Group
    )

  # -------------------------------------------------------
  # FINAL TRANSITION RESULTS
  # -------------------------------------------------------

  transition_results <- paired_events %>%
    mutate(
      Date = as.Date(Event_Timestamp),
      Time = format(Event_Timestamp, "%H:%M:%S"),
      Pair_Window = pair_window
    )

  # Save each dataset
  all_transition_results[[paste0("window_", pair_window)]] <- transition_results

  # -------------------------------------------------------
  # SUMMARY METRICS
  # -------------------------------------------------------

  temp_summary <- transition_results %>%
    summarise(
      Pair_Window = pair_window,
      Total_Events = n(),
      Mean_Time_Diff = mean(time_diff, na.rm = TRUE),
      Median_Time_Diff = median(time_diff, na.rm = TRUE)
    )

  summary_results <- bind_rows(summary_results, temp_summary)

  cat("Total Paired Events:",
      nrow(transition_results), "\n")

}

# =========================================================
# COMBINED RESULTS
# =========================================================

combined_transition_results <- bind_rows(all_transition_results)

# =========================================================
# VIEW SUMMARY TABLE
# =========================================================

print(summary_results)

# =========================================================
# PLOT WINDOW VS TOTAL EVENTS
# =========================================================
ggplot(summary_results,
       aes(x = Pair_Window,
           y = Total_Events)) +

  geom_line(linewidth = 1) +

  geom_point(size = 2) +

  # Labels for each point
  geom_text(
    aes(label = paste0("(", Pair_Window, ", ", Total_Events, ")")),
    vjust = -0.8,
    size = 4
  ) +

  labs(
    x = "Pair Window (seconds)",
    y = "Total Paired Events"
  ) +

  theme_minimal(base_size = 16)
```

Select which bats for data visualization
```{r}
library(ggplot2)
library(dplyr)

# Filter Sept 14-21 and count events per BAT_ID
plot_data <- transition_results %>%
  filter(Date >= as.Date("2025-09-14") & Date <= as.Date("2025-09-21")) %>%
  group_by(BAT_ID) %>%
  summarise(Total_Events = n(), .groups = "drop") %>%
  mutate(BAT_ID = factor(BAT_ID))

# Plot
ggplot(plot_data, aes(x = BAT_ID, y = Total_Events)) +
  geom_bar(stat = "identity", fill = "#4E79A7", color = "white", width = 0.7) +
  geom_text(aes(label = Total_Events), vjust = -0.4, size = 3.5, color = "gray30") +
  labs(
    title = "Total Transition Events per Bat (Sept 14-21)",
    x = "Bat ID",
    y = "Total Events"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )
```