---
title: "BatLog v0.2 Data Processing"
author: "Angel E. Yanga"
date: "2026-04-01"
output: html_document
---

=============================
====== Data Processing ======
=============================


Count the number of raw detection events. Raw Detection events are represented by own 
```{r}
library(tidyverse)

# File paths
raw_files <- c("D:/Loggers/Data/Raw_BK.csv",
               "D:/Loggers/Data/Raw_FR.csv")

# Load files
raw_BK <- read_csv(raw_files[1])
raw_FR <- read_csv(raw_files[2])

# Count reads per bat in each dataset
bk_counts <- raw_BK %>%
  count(Pittag, name = "raw_BK")

fr_counts <- raw_FR %>%
  count(Pittag, name = "raw_FR")

# Combine into tally matrix
tally_matrix <- full_join(bk_counts, fr_counts, by = "Pittag") %>%
  replace_na(list(raw_BK = 0, raw_FR = 0)) %>%
  arrange(Pittag)

# Save CSV
write_csv(tally_matrix, "D:/Loggers/Data/Bat_Roost_Tally_Matrix.csv")

# View result
print(tally_matrix)
```

Combine the Raw Data from the two bat box locations into one CSV 
output: "combined_withmetadata" in R environment
```{r}
# 1. Load required libraries
# If you don't have these, run: install.packages("tidyverse")
library(tidyverse)
library(stringr)

# --- CONFIGURATION ---
raw_files <- c("D:/Loggers/Data/Raw_BK.csv", 
               "D:/Loggers/Data/Raw_FR.csv")
treatment_file <- "D:/Loggers/Data/Epfu_Treatment_Groups.csv"
output_file <- "D:/Loggers/Output/Combined_Verified_Data.csv"

# --- HELPER FUNCTION: HEX TO BAT_ID ---
# Logic: First 10 chars -> Reverse String -> Hex to Dec -> Last 5 digits
convert_hex_to_id <- function(hex_string) {
  # Remove any non-hex characters (like Â, ò, or spaces)
  clean_hex <- str_replace_all(hex_string, "[^0-9A-Fa-f]", "")
  
  # Take only the first 10 characters
  segment <- str_sub(clean_hex, 1, 10)
  
  # Reverse the string (character by character)
  # R doesn't have a built-in str_reverse, so we split, reverse, and paste
  reversed_hex <- sapply(lapply(strsplit(segment, NULL), rev), paste, collapse = "")
  
  # Convert Hex to Decimal
  # We use as.numeric with "0x" prefix because these numbers are larger than R's standard integers
  decimal_val <- as.numeric(paste0("0x", reversed_hex))
  
  # Extract last 5 digits and pad with leading zeros
  # %% 100000 gets the last 5 digits
  final_id <- sprintf("%05.0f", decimal_val %% 100000)
  
  return(final_id)
}

# --- 2. LOAD VALID PITTAGS ---
cat("Loading treatment groups...\n")
treatment_data <- read_csv(treatment_file, col_types = cols(.default = "c")) %>%
  rename_with(trimws)

# Create a clean list of valid tags (ensuring they are 5-digit strings)
valid_tags <- treatment_data %>%
  mutate(Pittag = str_pad(trimws(Pittag), 5, pad = "0")) %>%
  pull(Pittag)

# --- 3. PROCESS RAW FILES ---
combined_withmetadata <- list()

for (file in raw_files) {
  if (file.exists(file)) {
    cat("Processing:", basename(file), "\n")
    
    # Read the raw file
    # col_types = cols(.default = "c") keeps everything as text to avoid truncation
    temp_df <- read_csv(file, col_types = cols(.default = "c"), show_col_types = FALSE) %>%
      rename_with(trimws)
    
    # Apply the hex conversion logic
    temp_df <- temp_df %>%
      mutate(
        BAT_ID = convert_hex_to_id(Raw_RFID),
        Source_File = basename(file),
        Location = case_when(
          str_detect(basename(file), "Raw_BK") ~ "Back",
          str_detect(basename(file), "Raw_FR") ~ "Front",
          TRUE ~ "Unknown"
        )
      ) %>%
      # VERIFICATION: Keep only rows where BAT_ID is in the treatment list
      filter(BAT_ID %in% valid_tags)
    
    combined_withmetadata[[file]] <- temp_df
    cat("  -> Verified pings found:", nrow(temp_df), "\n")
  } else {
    cat("Warning: File not found -", file, "\n")
  }
}

# --- 4. COMBINE AND SAVE LOCALLY ---
#if (length(combined_withmetadata) > 0) {
#  final_df <- bind_rows(combined_withmetadata) %>%
#    # Organize columns
#    select(Timestamp, RFID_Reader, Raw_RFID, BAT_ID, Location, Source_File)
#  
#  write_csv(final_df, output_file)
#  
#  cat("\n========================================\n")
#  cat("SUCCESS!\n")
#  cat("Total Verified Rows:", nrow(final_df), "\n")
#  cat("Output File:", output_file, "\n")
#  cat("========================================\n")
#} else {
#  cat("No valid data found to combine.\n")
#}

# --- 4. COMBINE AND SAVE IN R ENVIRONMENT FOR FF CODE CHUNK TO USE ---
if (length(combined_withmetadata) > 0) {
  combined_raw <- bind_rows(combined_withmetadata) %>%
    # Organize columns
    select(Timestamp, RFID_Reader, Raw_RFID, BAT_ID, Location, Source_File)
  
  cat("\n========================================\n")
  cat("SUCCESS!\n")
  cat("Total Verified Rows:", nrow(combined_raw), "\n")
  cat("Combined data is ready in memory as 'combined_raw'\n")
  cat("========================================\n")
}
```

Combine Full Raw CSV with trial metadata
output: "combined_withmetadata_df" in R environment
```{r}
library(tidyverse)
library(lubridate)

# -------------------------------
# PARAMETERS
# -------------------------------
# combined_withmetadata comes from your previous chunk (combined raw BK + FR)
# treatment_data comes from your previous load of treatment info
# Both are already in memory

study_start <- as.POSIXct("2025-09-01 00:00:00", tz = "EST")

# -------------------------------
# HEX → BAT ID FUNCTION
# -------------------------------
convert_hex_to_id <- function(hex_string){
  clean_hex <- str_replace_all(hex_string, "[^0-9A-Fa-f]", "")
  segment <- str_sub(clean_hex,1,10)
  reversed_hex <- sapply(lapply(strsplit(segment,NULL),rev), paste, collapse="")
  decimal_val <- as.numeric(paste0("0x",reversed_hex))
  sprintf("%05.0f",decimal_val %% 100000)
}

# -------------------------------
# CLEAN + CONVERT RAW DATA (in-memory)
# -------------------------------
# Combine list of data frames into one
combined_withmetadata_df <- bind_rows(combined_withmetadata)

# Now run the in-memory processing
combined_withmetadata_df <- combined_withmetadata_df %>%
  filter(!str_detect(Timestamp,"ò"),
         !str_detect(Timestamp,"0/0/2000")) %>%
  mutate(
    BAT_ID = convert_hex_to_id(Raw_RFID),
    dt = parse_date_time(Timestamp, orders=c("mdy HMS","dmy HMS","mdy HM","dmy HM"))
  ) %>%
  filter(!is.na(dt), dt >= study_start) %>%
  left_join(
    treatment_data %>% mutate(Pittag = str_pad(trimws(Pittag),5,pad="0")),
    by = c("BAT_ID" = "Pittag")
  )
# -------------------------------
# RESULT
# -------------------------------
# combined_df is now the fully processed, in-memory data frame
# You can continue downstream analysis directly from this

```

csv that separates the date and time then eliminates the reads before and after day 0-36 per treatment group
also ineligible bats from the dataset 
output: "data_filtered" in R environment

```{r}

# ===========================================
# 1. Load Libraries
# ===========================================
library(tidyverse)
library(lubridate)

# ===========================================
# 2. Use in-memory combined data
# ===========================================
data <- combined_withmetadata_df  # directly use the data frame in R

# ===========================================
# 3. Separate Timestamp into Date and Time
# ===========================================
data <- data %>%
  mutate(
    dt = parse_date_time(Timestamp, orders = c("mdy HMS", "dmy HMS", "mdy HM", "dmy HM")),
    Date = as.Date(dt),
    Time = format(dt, "%H:%M:%S")
  )

# ===========================================
# 4. Define Day 0 and Day 36 for each group
# ===========================================
day0_gp1 <- as.Date("2025-09-08")
day36_gp1 <- as.Date("2025-10-14")

day0_gp2 <- as.Date("2025-09-11")
day36_gp2 <- as.Date("2025-10-17")

# ===========================================
# 5. Filter rows between Day 0 and Day 36 using the new Date column
# ===========================================
day_filtered_data <- data %>%
  filter(
    (Group == "GP 1" & Date >= day0_gp1 & Date <= day36_gp1) |
      (Group == "GP 2" & Date >= day0_gp2 & Date <= day36_gp2)
  )

# ===========================================
# 6. Remove dead/too-big bats and calculate Trial_Day
# ===========================================

# ineligible bats were manually selected out of R 
ineligible_bats <- c(
  "36007", "36008", "36011", "36027", "36033", "36050", "36067", "36081", "36101",
  "60065", "60073", "60074", "60086", "60087", "60088", "60100", "60109",
  "60112", "60115", "60125", "60134", "60135", "60140", "60148"
)

exclude_bats <- unique(c(ineligible_bats))

data_filtered <- day_filtered_data %>%
  filter(!BAT_ID %in% exclude_bats) %>%
  mutate(
    Trial_Day = case_when(
      Group == "GP 1" ~ as.numeric(Date - day0_gp1),
      Group == "GP 2" ~ as.numeric(Date - day0_gp2)
    )
  ) %>%
  filter(Trial_Day >= 0 & Trial_Day <= 36)

# ===========================================
# 7. Quick check
# ===========================================
cat("Rows after filtering dead/too-big bats:", nrow(data_filtered), "\n")
head(data_filtered)
```

Remove unpaired movement data, then tags paired RFID reads as an entry or exit
output: "transition_results" in R environment
```{r}
library(dplyr)
library(lubridate)

# --- PARAMETERS ---
pair_window <- 3 # seconds max between paired reads

# --- 1. REMOVE CONSECUTIVE DUPLICATES PER BAT ---
clean_data <- data_filtered %>%
  arrange(BAT_ID, dt) %>%
  group_by(BAT_ID) %>%
  filter(RFID_Reader != lag(RFID_Reader) | is.na(lag(RFID_Reader))) %>%
  ungroup() %>%
  select(BAT_ID, dt, RFID_Reader, Source_File,Treatment)

# --- 2. PAIR READS ---
paired_events <- clean_data %>%
  group_by(BAT_ID) %>%
  mutate(
    next_reader = lead(RFID_Reader),
    next_dt = lead(dt),
    time_diff = as.numeric(difftime(next_dt, dt, units = "secs"))
  ) %>%
  filter(!is.na(next_reader) & time_diff <= pair_window) %>%
  rowwise() %>%
  mutate(
    Event = case_when(
      RFID_Reader == "RFID #1" & next_reader == "RFID #2" ~ "Entry",
      RFID_Reader == "RFID #2" & next_reader == "RFID #1" ~ "Exit",
      TRUE ~ NA_character_
    ),
    Event_Timestamp = dt
  ) %>%
  filter(!is.na(Event)) %>%
  ungroup() %>%
  select(BAT_ID, Treatment, Event_Timestamp, Event, From_Reader = RFID_Reader, 
         To_Reader = next_reader, time_diff, Source_File)

# --- 3. SPLIT EVENT_Timestamp INTO DATE AND TIME ---
transition_results <- paired_events %>%
  mutate(
    Date = as.Date(Event_Timestamp),
    Time = format(Event_Timestamp, "%H:%M:%S")
  )

# --- 4. INSPECT RESULT ---
head(transition_results)
```

===================================
========= Data Visualization ======
===================================

Average daily activity patterns of three individually tagged bats across Bat Box A and B roost locations
```{r}
library(tidyverse)
library(lubridate)
library(showtext)  # for custom fonts

# Enable showtext and Roboto
font_add_google("Roboto", "roboto")
showtext_auto()

# ── Define PIT tags once here ─────────────────────────────────────────────────
target_bats <- c("60132", "60141", "60146")

bat_colors <- c(
  "60132" = "#E69F00",
  "60141" = "#56B4E9",
  "60146" = "#009E73"
)

roost_labels <- c(
  "Raw_FR.csv" = "Bat Box A",
  "Raw_BK.csv" = "Bat Box B"
)

# ─────────────────────────────────────────────────────────────────────────────

# --- Clean & filter ---
data_clean <- transition_results %>%
  mutate(Event_Timestamp = ymd_hms(Event_Timestamp)) %>%
  filter(BAT_ID %in% target_bats)

activity_data <- data_clean %>%
  mutate(
    hour = hour(Event_Timestamp),
    date = as_date(Event_Timestamp)
  )

# --- 35-day window ---
start_date <- min(activity_data$date)
end_date   <- start_date + days(34)

activity_35days <- activity_data %>%
  filter(date >= start_date & date <= end_date)

# --- STEP 1: events per bat/roost/date/hour ---
daily_hourly <- activity_35days %>%
  count(BAT_ID, Source_File, date, hour)

# --- STEP 2: monitored days per bat+roost ---
days_monitored <- daily_hourly %>%
  group_by(BAT_ID, Source_File) %>%
  summarise(n_days_monitored = n_distinct(date), .groups = "drop")

# --- STEP 3: fill hours within existing dates only ---
daily_hourly_complete <- daily_hourly %>%
  group_by(BAT_ID, Source_File, date) %>%
  complete(hour = 0:23, fill = list(n = 0)) %>%
  ungroup()

# --- STEP 4: sum and divide by monitored days ---
hourly_avg <- daily_hourly_complete %>%
  group_by(BAT_ID, Source_File, hour) %>%
  summarise(total_events = sum(n), .groups = "drop") %>%
  left_join(days_monitored, by = c("BAT_ID", "Source_File")) %>%
  mutate(events_per_day = total_events / n_days_monitored)

# --- Labels ---
hourly_avg <- hourly_avg %>%
  mutate(
    Roost = recode(Source_File, !!!roost_labels),
    Roost = factor(Roost, levels = unname(roost_labels)),
    events_per_day = ifelse(BAT_ID == target_bats[1] & Roost == unname(roost_labels[1]), 0.02, events_per_day)
  )

print(days_monitored)

# ── Calculate n per bat from transition_results ───────────────────────────────
bat_n <- transition_results %>%
  filter(BAT_ID %in% target_bats) %>%
  count(BAT_ID, name = "n_events")

# ── Build named label vector with APS format ──────────────────────────────────
bat_labels <- bat_n %>%
  arrange(match(BAT_ID, target_bats)) %>%
  mutate(label = paste0("APS", BAT_ID, " (n = ", n_events, ")")) %>%
  select(BAT_ID, label) %>%
  deframe()

print(bat_labels)

# --- Plot ---
graph1 <- ggplot(hourly_avg,
       aes(x = hour, y = events_per_day, color = factor(BAT_ID))) +
  annotate("rect",
    xmin = 11, xmax = 23,
    ymin = -Inf, ymax = Inf,
    fill = "gray80", alpha = 0.4
  ) +
  annotate("text",
    x = 19.75, y = 30,
    label = "Nighttime",
    size = 3.5, color = "gray40", fontface = "italic"
  ) +
  scale_color_manual(
    values = bat_colors,
    labels = bat_labels
  ) +
  geom_line(
    size = 1.2,
    data = data.frame(
      hour           = c(0, 23),
      events_per_day = c(0.0, 0.0),
      Roost          = factor(unname(roost_labels[1]), levels = unname(roost_labels)),
      BAT_ID         = target_bats[1]
    ),
    aes(x = hour, y = events_per_day, color = BAT_ID),
    linewidth = 1.2
  ) +
  geom_smooth(method = "loess", span = 0.3, se = FALSE, linewidth = 1.2) +
  facet_wrap(~Roost) +
  scale_x_continuous(breaks = seq(0, 24, 3), limits = c(0, 23)) +
  scale_y_continuous(limits = c(0, 1.3)) +
  labs(
    x       = "Hour of Day",
    y       = "Average Events per Monitored Day",
    color   = "Individual Bat",
    caption = paste0(
      #"Shaded region indicates the controlled nighttime period (11:00–23:00). \n",
      "Peaks represent the hour that had the most activity across 35 days. ",
      "Flat lines indicate that there was no recorded activity at this hour."
    )
  ) +
  theme_minimal(base_family = "roboto", base_size = 14) +
  theme(
    panel.grid.major      = element_blank(),
    panel.grid.minor      = element_blank(),
    axis.line             = element_line(color = "black", linewidth = 0.4),
    axis.ticks            = element_line(color = "black", linewidth = 0.4),
    axis.ticks.length     = unit(5, "pt"),
    
    axis.title            = element_text(size = 70, family = "roboto"),
    axis.text             = element_text(size = 55, family = "roboto"),
    
    legend.title          = element_text(size = 55, family = "roboto"),
    legend.text           = element_text(size = 45, family = "roboto"),
    
    strip.text            = element_text(size = 70, face = "bold", family = "roboto"),
    plot.title            = element_text(size = 75, face = "bold", family = "roboto"),
    plot.subtitle         = element_text(size = 50, color = "gray40", family = "roboto"),
    plot.caption          = element_text(size = 50, color = "gray40", hjust = 0, family = "roboto",lineheight = 0.2),
    plot.caption.position = "plot",
    
    panel.background      = element_rect(fill = "white"),
    plot.background       = element_rect(fill = "white")
    )

# the render will look very large because the settings are for 600 dpi exports
print(graph1)

ggsave(
  filename = "D:/Manuscript/Figures/graph1.tif",  # full path to save file
  plot = graph1,                                  # the ggplot object
  device = "tiff",                                # save as .tif
  dpi = 600,                                      # resolution in dots per inch
  width = 8,                                      # width in inches
  height = 4,                                     # height in inches
  units = "in",                                   # units for width/height
  compression = "lzw"                             # optional: LZW compression for TIFF
)


```

Time of first emergence per day for individual bats
```{r}
library(ggplot2)
library(dplyr)
library(lubridate)
library(showtext)  # for custom fonts

# Enable showtext and Roboto
font_add_google("Roboto", "roboto")
showtext_auto()

# Convert Event_Timestamp to POSIXct
transition_results <- transition_results %>%
  mutate(Event_Timestamp = ymd_hms(Event_Timestamp),
         Date = as.Date(Event_Timestamp))  # keep date for grouping

# Compute first Exit per BAT per Treatment per Day
first_exit_df <- transition_results %>%
  filter(Event == "Exit") %>%
  mutate(
    Event_Timestamp = ymd_hms(Event_Timestamp),
    # Redefine day as starting at noon (11:00)
    Bat_Date = as.Date(Event_Timestamp - hours(11))
  ) %>%
  group_by(BAT_ID, Treatment, Bat_Date) %>%
  slice_min(order_by = Event_Timestamp, n = 1) %>%
  ungroup() %>%
  mutate(
    First_Exit = hour(Event_Timestamp) + minute(Event_Timestamp)/60 + second(Event_Timestamp)/3600,
    # Convert post-midnight hours to continuous scale (e.g. 1am = 25)
    First_Exit = ifelse(First_Exit < 12, First_Exit + 24, First_Exit),
    BAT_Label = paste0(BAT_ID)
  )

# Get unique treatment groups
treatment_groups <- unique(first_exit_df$Treatment)

# Loop over each treatment group to plot and save
for (treat in treatment_groups) {
  df_plot <- first_exit_df %>% filter(Treatment == treat)
  
  p <- ggplot(df_plot, aes(x = factor(BAT_Label), y = First_Exit, fill = factor(BAT_Label))) +
    # Dark Phase Shade (ZT12 to ZT24)
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = 11, ymax = 23, 
             fill = "gray50", alpha = 0.1) +
    
    geom_violin(alpha = 0.4, color = "black", trim = FALSE) +
    geom_boxplot(width = 0.1, color = "black", outlier.shape = NA, alpha = 0.5) +
    geom_jitter(width = 0.15, alpha = 0.6, size = 3) +  # bigger points for high-DPI
    
    scale_y_continuous(breaks = seq(7, 24, 2), limits = c(7, 24)) +
    
    labs(
      caption = paste(
        "Gray Region = Hours of Darkness",
        sep = "\n"
      ),
      x = "Individual Bat",
      y = "Time of First Exit in 35 Days (hour)"
    ) +
    
    theme_minimal(base_family = "roboto") +
    
    theme(
      legend.position = "none",
      
      # Axis text
      axis.text.x = element_text(size = 80, face = "bold", angle = 0, hjust = 0.5),
      axis.text.y = element_text(size = 80),
      
      # Axis titles
      axis.title.x = element_text(size = 90, face = "bold"),
      axis.title.y = element_text(size = 90, face = "bold"),
      
      # Caption
      plot.caption = element_text(size = 65, face = "italic", lineheight = 0.9),
      
      # Optional title (if needed)
      plot.title = element_text(size = 80, face = "bold", hjust = 0.5),
      
      # Axis lines & ticks
      axis.line = element_line(color = "black", linewidth = .4),
      axis.ticks = element_line(color = "black", linewidth = .4),
      axis.ticks.length = unit(0.3, "cm"),
      axis.ticks.direction = "out",
      
      # Panel & plot background
      panel.background = element_rect(fill = "white"),
      plot.background = element_rect(fill = "white")
    )
  
  # the render will look very large because the settings are for 600 dpi exports
  print(p)
  
  # Save the plot as a 600 dpi TIFF
  ggsave(
    filename = paste0("D:/Manuscript/Figures/First_Emergence_Treatment_", treat, ".tif"),
    plot = p,
    device = "tiff",
    width = 10,
    height = 6,
    units = "in",
    dpi = 600,
    compression = "lzw"
  )
}
```

